% !TEX TS-program = xelatex

\documentclass[11pt]{article}

\input{../shared/preamble.tex}

\renewcommand{\fragmentnumber}{Fragment 1}
\renewcommand{\fragmenttitle}{Fused determiner-heads}
\renewcommand{\fragmentdate}{2026-05-07}

\newcommand{\GenNP}{\Cat{GenNP}}
\newcommand{\SynDmr}{\mathrm{SynDeterminer}}
\newcommand{\InterpAs}{\mathrm{InterpretsAs}}
\newcommand{\FusedDetHead}{\mathrm{FusedDetHead}}
\newcommand{\FusedDetHeadNP}{\mathrm{FusedDetHeadNP}}
\newcommand{\CanonicalDetHeadNP}{\mathrm{CanonicalDetHeadNP}}

\begin{document}
\fragmenthead

\noindent
This fragment tests the local end of the project: can a CGEL-style category /
function distinction be stated as a small constraint on labelled trees? The
target is the familiar contrast between an NP with separate daughters in
Determiner and Head functions, and an NP in which one determiner realizes both
functions.

\section*{1.~Phenomenon}

CGEL treats examples such as \mention{three}, \mention{some}, and
\mention{many} in independent NP use as cases of fused determiner-heads. The
fusion is stated over functions: the same daughter fills Determiner and Head.
The fused determiner is a DP in the ordinary simple cases; in genitive cases it
can be a genitive NP.

\ea\label{ex:fused-three} \textit{Pat invited two students; I invited three.}\z

\ea\label{ex:fused-many} \textit{Some of the reports were late, but many arrived on time.}\z

\ea\label{ex:fused-genitive} \textit{Kim's proposal was stronger than Pat's.}\z

The descriptive problem is small but useful. In \mention{two students},
\mention{two} is the determiner, and the nominal \mention{students} fills Head.
In independent \mention{three}, the determiner fills both functions. In
\mention{Pat's}, the determiner is a genitive NP. A fragment should therefore
distinguish ordinary determiner-plus-Head NPs from fused determiner-head NPs
without recasting the fused function pair as either a lexical category claim or
a semantic category claim.

\section*{2.~Vocabulary}

Use the shared relations $\imdom$ and $\dm$, and the function relations
$\Hd$ and $\Det$. Add one local syntactic predicate for determiners in this
draft:

\ea\label{ex:determiner-predicate}
$\SynDmr(x) \defas \DP(x) \vee \GenNP(x)$\\
\gloss{$x$ is a syntactic determiner: a DP or a genitive NP in this draft}
\z

The predicate $\GenNP$ is intentionally provisional: a fuller fragment should
say whether it is a feature-bearing subtype of $\NP$, a separate abbreviation,
or a definable configuration. For now it keeps the category/function issue
visible.

No semantic predicate is introduced here. In particular, $\SynDmr(x)$ does not
say that $x$ denotes a generalized quantifier, a choice function, a possessive
relation, or any other semantic category. If a later fragment needs that
bridge, it should add a separate interpretation relation, such as
$\InterpAs(x,s)$, with explicit warrants.

\section*{3.~Specification}

First define the fused function-pair relation:

\ea\label{ex:fused-det-head}
$\begin{aligned}
\FusedDetHead(d,n) \defas {} & n \imdom d \wedge \Det(n,d) \wedge \Hd(n,d).
\end{aligned}$\\
\gloss{$d$ is both Determiner and Head of $n$}
\z

An ordinary NP with separate Determiner and nominal Head daughters can be
characterised as follows:

\ea\label{ex:canonical-det-head}
$\begin{aligned}
\CanonicalDetHeadNP(n) \defas {} & \NP(n) \wedge
  (\exists d)(\exists h)[n \imdom d \wedge n \imdom h \wedge d \neq h \\
& \qquad {} \wedge \Det(n,d) \wedge \SynDmr(d)
  \wedge \Hd(n,h) \wedge \Nom(h)].
\end{aligned}$\\
\gloss{an NP with a determiner and a distinct nominal Head}
\z

The fused determiner-head configuration is the corresponding case where one
determiner realizes both functions and no separate nominal Head daughter is
present:

\ea\label{ex:fused-det-head-np}
$\begin{aligned}
\FusedDetHeadNP(n) \defas {} & \NP(n) \wedge
  (\exists d)[\SynDmr(d) \wedge \FusedDetHead(d,n) \\
& \qquad {} \wedge
  \neg(\exists h)[n \imdom h \wedge h \neq d \wedge \Hd(n,h) \wedge \Nom(h)]].
\end{aligned}$\\
\gloss{an NP whose determiner is both Determiner and Head}
\z

For simple cases such as \mention{three}, this is a first-order local
definition. Partitive cases add one more daughter, but the fused determiner is
still the DP in the fused function pair:

\ea\label{ex:partitive-fused}
$\begin{aligned}
\mathrm{PartitiveFusedDetHeadNP}(n) \defas {} & \FusedDetHeadNP(n) \wedge
  (\exists p)[n \imdom p \wedge \Cmp(n,p) \wedge \PP(p)].
\end{aligned}$\\
\gloss{a fused determiner-head NP with an \mention{of}-PP complement}
\z

Nothing in the formula assigns the fused element to a semantic category. It
says only that this tree has one daughter bearing two syntactic functions. In
the ordinary cases that daughter is a DP.

\section*{4.~Worked instance}

Take \mention{I invited three}. Let $n$ be the object NP and $d$ the DP
\mention{three}. The intended tree makes $d$ a determiner and gives $d$ both
the Determiner and Head functions in $n$. There is no distinct nominal Head
daughter, so $n$ satisfies $\FusedDetHeadNP(n)$.

The contrast with \mention{two students} is immediate. In that NP, the
determiner \mention{two} fills $\Det(n,d)$, while the nominal
\mention{students} fills $\Hd(n,h)$. That NP satisfies
$\CanonicalDetHeadNP(n)$, not $\FusedDetHeadNP(n)$.

The genitive case shows why the analysis cannot be stated as a lexical category
predicate. In \mention{Kim's proposal was stronger than Pat's}, the fused
daughter in the second comparison is a genitive NP. It still fills the
Determiner and Head functions together. Any claim about its possessive
semantics would be an additional bridge claim, not part of
$\FusedDetHeadNP(n)$.

\section*{5.~Open questions}

\begin{enumerate}[label=(\alph*),leftmargin=2em,itemsep=0.25em]
  \item The inventory of syntactic determiner types needs checking. DP and
        genitive NP are included here because they expose the main
        category/function contrast, but CGEL's full set of determiners may need
        a richer definition.
  \item Semantic bridges are deliberately absent. Claims about quantification,
        comparison, partitivity, or possessive interpretation need separate
        semantic vocabulary plus bridge conditions.
  \item Partitives probably require tighter treatment of the \mention{of}-PP:
        it is not just any complement PP.
  \item The inventory of non-fused determiners should be kept separate from the
        fused-function configuration unless the interaction is
        explicitly modelled.
  \item CGEL page references and the Payne--Huddleston--Pullum treatment of
        fused functions need to be checked before this becomes a citable
        fragment.
\end{enumerate}

\end{document}
