% !TEX TS-program = xelatex

\documentclass[11pt]{article}

\input{../shared/preamble.tex}

\renewcommand{\fragmentnumber}{Fragment 1}
\renewcommand{\fragmenttitle}{Determiner-head fusion with determinatives}
\renewcommand{\fragmentdate}{2026-05-07}

\newcommand{\Dtr}{\Cat{Dtr}}
\newcommand{\Fused}{\mathrm{Fused}}
\newcommand{\FusedDetHeadNP}{\mathrm{FusedDetHeadNP}}
\newcommand{\CanonicalDetHeadNP}{\mathrm{CanonicalDetHeadNP}}

\begin{document}
\fragmenthead

\noindent
This fragment tests the local end of the project: can a CGEL-style category /
function distinction be stated as a small constraint on labelled trees? The
target is the familiar contrast between an NP with a separate determiner and
nominal head, and an NP in which one determinative-category node realizes both
the determiner function and the Head function.

\section*{1.~Phenomenon}

CGEL treats examples such as \mention{three}, \mention{some}, and
\mention{many} in independent NP use as cases of determiner-head fusion:
a determinative is not followed by an overt nominal head, and the same
determinative-category node fills the determiner function and the Head
function.

\ea
\textit{Pat invited two students; I invited three.}
\z

\ea
\textit{Some of the reports were late, but many arrived on time.}
\z

\ea
\textit{These are useful; those are defective.}
\z

The descriptive problem is small but useful. In \mention{two students}, the
determinative \mention{two} fills the determiner function (Det), and the
nominal \mention{students} fills Head. In independent \mention{three}, a
single determinative-category item fills both functions. A fragment should
therefore distinguish ordinary determiner-plus-Head NPs from fused
determiner-head NPs without confusing the category \term{determinative} with
the function \term{determiner}.

\section*{2.~Vocabulary}

Use the shared relations $\imdom$ and $\dm$, and the function relations
$\Hd$ and $\Det$. Add one local lexical-category predicate:

\ea
$\Dtr(x)$: $x$ is a determinative-category node.
\z

For this draft, $\Fused(d,n)$ is a bookkeeping predicate saying that the
determinative node $d$ inherits whatever interpretation would ordinarily be
supplied by the missing nominal material in NP $n$. The point is not yet to
formalise recoverability; it is to isolate the tree configuration that calls
for it.

\section*{3.~Specification}

An ordinary NP with a separate determiner and nominal Head can be characterised
as follows:

\ea
$\begin{aligned}
\CanonicalDetHeadNP(n) \defas {} & \NP(n) \wedge
  (\exists d)(\exists h)[n \imdom d \wedge n \imdom h \wedge d \neq h \\
& \qquad {} \wedge \Det(n,d) \wedge \Dtr(d) \wedge \Hd(n,h) \wedge \Nom(h)].
\end{aligned}$\\
\gloss{an NP with a determinative in determiner function and a distinct nominal Head}
\z

The fused determiner-head configuration is the corresponding case where one
determinative-category daughter realizes both functions and no separate
nominal Head daughter is present:

\ea
$\begin{aligned}
\FusedDetHeadNP(n) \defas {} & \NP(n) \wedge
  (\exists d)[n \imdom d \wedge \Det(n,d) \wedge \Hd(n,d)
    \wedge \Dtr(d) \wedge \Fused(d,n) \\
& \qquad {} \wedge
  \neg(\exists h)[n \imdom h \wedge h \neq d \wedge \Hd(n,h) \wedge \Nom(h)]].
\end{aligned}$\\
\gloss{an NP whose determinative daughter is both determiner and Head}
\z

For simple cases such as \mention{three}, this is a first-order local
definition. Partitive cases add one more daughter:

\ea
$\begin{aligned}
\mathrm{PartitiveFusedDetHeadNP}(n) \defas {} & \FusedDetHeadNP(n) \wedge
  (\exists p)[n \imdom p \wedge \Cmp(n,p) \wedge \PP(p)].
\end{aligned}$\\
\gloss{a fused determiner-head NP with an \mention{of}-PP complement}
\z

Nothing in the formula says that determinatives are nouns, pronouns, or a
separate primary category. It says only that this tree has a determinative
node simultaneously in determiner and Head functions.

\section*{4.~Worked instance}

Take \mention{I invited three}. Let $n$ be the object NP and $d$ the lexical
node \mention{three}. In the intended tree, $n \imdom d$, $\Det(n,d)$,
$\Hd(n,d)$, and $\Dtr(d)$. There is no distinct nominal Head daughter.
Therefore $n$ satisfies $\FusedDetHeadNP(n)$.

The contrast with \mention{two students} is immediate. In that NP, the
determinative \mention{two} fills $\Det(n,d)$, while the nominal
\mention{students} fills $\Hd(n,h)$. That NP satisfies
$\CanonicalDetHeadNP(n)$, not $\FusedDetHeadNP(n)$.

\section*{5.~Open questions}

\begin{enumerate}[label=(\alph*),leftmargin=2em,itemsep=0.25em]
  \item This draft treats $\Fused$ as primitive. A fuller fragment needs a
        recoverability condition for the missing nominal contribution, or a
        decision that the simple cases should be analysed as ordinary
        determinatives in Head function without fusion.
  \item Partitives probably require tighter treatment of the \mention{of}-PP:
        it is not just any complement PP.
  \item The defective articles \mention{the} and \mention{a} should be kept
        outside the productive simple-head generalisation unless their special
        status is explicitly modelled.
  \item CGEL page references and the Payne--Huddleston--Pullum treatment of
        fused functions need to be checked before this becomes a citable
        fragment.
\end{enumerate}

\end{document}
