% !TEX TS-program = xelatex

\documentclass[11pt]{article}

\input{../shared/preamble.tex}

\renewcommand{\fragmentnumber}{Fragment 2}
\renewcommand{\fragmenttitle}{Independent relative \mention{whose}}
\renewcommand{\fragmentdate}{2026-05-07}

\newcommand{\Vacancy}{\mathrm{Vacancy}}
\newcommand{\NPVac}{\mathrm{NPVacancy}}
\newcommand{\Bridge}{\mathrm{Bridge}}
\newcommand{\DomOrdered}{\mathrm{DomOrdered}}
\newcommand{\DomContinuous}{\mathrm{DomContinuous}}
\newcommand{\DomChain}{\mathrm{DomChain}}
\newcommand{\BridgeLinked}{\mathrm{BridgeLinked}}
\newcommand{\Root}{\mathrm{Root}}
\newcommand{\NPBindable}{\mathrm{NPBindable}}
\newcommand{\RelWhose}{\mathrm{RelWhose}}
\newcommand{\PossRec}{\mathrm{PossessumRecoverable}}

\begin{document}
\fragmenthead

\noindent
This fragment tests whether Pullum's bridge-linked dominance-chain machinery
can be used for an empirically delicate CGEL construction: relative
\mention{whose} without an overt following possessum.

\section*{1.~Phenomenon}

Independent relative \mention{whose} is the relative-genitive construction in
which \mention{whose} appears without a following noun:

\ea
\textit{I was going to visit Lucy, a friend of whose had told us of the accident.}
\z

\ea
\textit{Norway, whose standard of living is the envy of the world, has thicker kids than Singapore, whose isn't.}
\z

\ea
\textit{The man whose these are hath gotten me with child.}
\z

The construction is rare and pragmatically constrained. The syntactic task for
this fragment is narrower: state the tree condition under which a relative
clause contains a relative-genitive phrase whose possessum is not overtly
realised, and connect that phrase to a vacancy or predicative gap in the
relative-clause nucleus.

\section*{2.~Vocabulary}

The definitions below adapt Pullum's handout notation. Immediate dominance is
$\imdom$; dominance is $\dm$. The relation $\Bridge(x,y)$ is initially limited
to Head and Complement links:

\ea
$\Bridge(x,y) \defas \Hd(x,y) \vee \Cmp(x,y)$\\
\gloss{$y$ is linked to $x$ by a bridge-preserving function}
\z

We add $\RelWhose(w)$ for a lexical relative-genitive \mention{whose} node and
$\PossRec(w)$ for the discourse condition that the missing possessum is
recoverable. The second predicate is not syntactic; it marks the point where
this construction is known to need information-structural licensing.

\section*{3.~Specification}

Pullum's basic vacancy predicate can be stated as:

\ea
$\Vacancy(x) \defas \Phrasal(x) \wedge \neg(\exists y)[x \imdom y]$\\
\gloss{a phrasal node with no children}
\z

\ea
$\NPVac(x) \defas \Vacancy(x) \wedge \NP(x)$\\
\gloss{an NP vacancy}
\z

The relevant chain definitions are:

\ea
$\begin{aligned}
\DomOrdered(X) \defas {} &
  (\forall x)(\forall y)[(X(x) \wedge X(y)) \Rightarrow
    (x \dm y \vee y \dm x)].
\end{aligned}$\\
\gloss{all nodes in $X$ are ordered by dominance}
\z

\ea
$\begin{aligned}
\DomContinuous(X) \defas {} &
  (\forall x)(\forall y)(\forall z)[(X(x) \wedge X(z) \wedge x \dm y
    \wedge y \dm z) \Rightarrow X(y)].
\end{aligned}$\\
\gloss{$X$ contains the intervening nodes on its dominance path}
\z

\ea
$\DomChain(X) \defas \DomOrdered(X) \wedge \DomContinuous(X)$\\
\gloss{a dominance chain}
\z

\ea
$\begin{aligned}
\BridgeLinked(X) \defas {} & \DomChain(X) \wedge
  (\forall x)(\forall y)[(X(x) \wedge X(y) \wedge x \imdom y)
    \Rightarrow \Bridge(x,y)].
\end{aligned}$\\
\gloss{a dominance chain whose immediate links are bridge links}
\z

\ea
$\Root(X,r) \defas (\forall x)[X(x) \Leftrightarrow r \dm x]$\\
\gloss{$X$ is the set of nodes dominated by root $r$}
\z

\ea
$\begin{aligned}
\NPBindable(r) \defas {} &
  (\exists X)[\Root(X,r) \wedge \BridgeLinked(X) \wedge
    (\exists z)[X(z) \wedge \NPVac(z)]].
\end{aligned}$\\
\gloss{$r$ roots a bridge-linked constituent containing an NP vacancy}
\z

For independent relative \mention{whose}, the draft adds a condition on the
relative clause:

\ea
$\begin{aligned}
\mathrm{IndRelWhoseClause}(c) \defas {} & \Cl(c) \wedge
  (\exists w)(\exists r)[\RelWhose(w) \wedge c \dm w \wedge c \dm r \\
& \qquad {} \wedge \NPBindable(r) \wedge \PossRec(w)].
\end{aligned}$\\
\gloss{a relative clause containing independent \mention{whose} and a bindable nucleus}
\z

\section*{4.~Worked instance}

In \mention{Singapore, whose isn't}, the relative clause contains a
relative-genitive \mention{whose} and a nucleus with stranded auxiliary
\mention{isn't}. The missing possessum is recoverable from the parallel
antecedent \mention{Norway, whose standard of living is ...}. In this use,
$\PossRec(w)$ is licensed by contrastive parallelism.

The syntactic part then asks whether the nucleus rooted at $r$ contains a
bridge-linked NP vacancy. If it does, $r$ satisfies $\NPBindable(r)$ and the
relative clause satisfies $\mathrm{IndRelWhoseClause}(c)$. The formula
therefore keeps two claims apart: the structural possibility of a
relative-genitive dependency, and the pragmatic recoverability of the
possessum.

\section*{5.~Open questions}

\begin{enumerate}[label=(\alph*),leftmargin=2em,itemsep=0.25em]
  \item Some attested cases are predicative-complement cases rather than
        simple NP-vacancy cases. Those may require a second vacancy type.
  \item Oblique genitive examples such as \mention{a friend of whose} may be
        locally licensed by the possessum noun inside the relative phrase,
        rather than by a distant vacancy in the nucleus.
  \item $\PossRec$ is a placeholder for discourse licensing, especially
        contrastive parallelism, deictic anchoring, and structural anchoring.
        The syntax fragment should say where that condition attaches, but not
        pretend that it is a tree-only predicate.
  \item CGEL page references are available in the separate \mention{whose}
        project but should be rechecked against the book before citation.
\end{enumerate}

\end{document}
