% !TEX TS-program = xelatex

\documentclass[11pt]{article}

\input{../shared/preamble.tex}

\renewcommand{\fragmentnumber}{Fragment 3}
\renewcommand{\fragmenttitle}{Across-the-board extraction}
\renewcommand{\fragmentdate}{2026-05-07}

\newcommand{\Vacancy}{\mathrm{Vacancy}}
\newcommand{\NPVac}{\mathrm{NPVacancy}}
\newcommand{\Bridge}{\mathrm{Bridge}}
\newcommand{\BridgeLinked}{\mathrm{BridgeLinked}}
\newcommand{\Root}{\mathrm{Root}}
\newcommand{\NPBindable}{\mathrm{NPBindable}}
\newcommand{\Coordination}{\Cat{Coordination}}
\newcommand{\Coordinate}{\mathrm{Coordinate}}
\newcommand{\Coordinator}{\Cat{Coordinator}}
\newcommand{\CoordNPBindable}{\mathrm{CoordNPBindable}}
\newcommand{\ATB}{\mathrm{ATB}}
\newcommand{\ParallelGap}{\mathrm{ParallelGap}}

\begin{document}
\fragmenthead

\noindent
This fragment is a deliberate stress test. Pullum's handout shows how a
SLASH-free, logic-on-trees treatment can describe ordinary unbounded
dependencies, then flags Across-the-Board extraction as the hard residue where
Gazdar's GPSG analysis is especially elegant.

\section*{1.~Phenomenon}

Across-the-Board extraction allows one preposed phrase to correspond to gaps
inside each coordinate of a coordination:

\ea
\textit{Those, I either sent back or paid for.}
\z

\ea
\ungram{\textit{Those, I either paid the bill or sent back.}}
\z

\ea
\ungram{\textit{Those, I either sent back or paid the bill.}}
\z

The good case has matching gaps in both coordinates. The bad cases have only
one gap: one coordinate is independently compatible with the extracted NP, but
the other is not. The fragment asks whether this can be stated directly over
trees, without reintroducing a SLASH-valued category.

\section*{2.~Vocabulary}

Reuse Pullum's vacancy and bindability machinery, but add CGEL-style
coordination vocabulary:

\ea
$\Coordination(c)$: $c$ is a coordination.\\
$\Coordinate(c,k)$: $k$ fills Coordinate function in coordination $c$.\\
$\Coordinator(m)$: $m$ is a coordinator.\\
$\Mk(k,m)$: $m$ fills Marker function inside coordinate $k$.\\
$\ParallelGap(z_1,z_2)$: vacancies $z_1$ and $z_2$ are syntactically parallel.
\z

A coordination may be unmarked, or the final Coordinate daughter may contain a
coordinator in Marker function. The Marker is not a dependent of the
coordination as a whole, and it is not itself a Coordinate. The ATB condition
below quantifies over Coordinate daughters, not over Marker dependents inside
them.

The predicate $\ParallelGap$ is a deliberate pressure point. For NP extraction,
it might mean that both vacancies are complements of comparable lexical heads,
or that both are linked to the same extracted prenucleus NP by the same
syntactic relation. It is not a semantic or thematic role predicate. If that
extra condition cannot be reduced to syntactic relations such as
function-labelled dominance plus a shared prenucleus relation, the fragment has
located a real bridge problem rather than solved it.

\section*{3.~Specification}

Use the same basic vacancy definitions:

\ea
$\Vacancy(x) \defas \Phrasal(x) \wedge \neg(\exists y)[x \imdom y]$
\z

\ea
$\NPVac(x) \defas \Vacancy(x) \wedge \NP(x)$
\z

Assume the Pullum-style definition of $\NPBindable(r)$: $r$ roots a
bridge-linked dominance chain containing an NP vacancy. Then define a
Coordinate daughter as NP-binding if it has such a root:

\ea
$\begin{aligned}
\CoordNPBindable(k) \defas {} &
  (\exists r)[k \dm r \wedge \NPBindable(r)].
\end{aligned}$\\
\gloss{Coordinate daughter $k$ contains an NP-bindable domain}
\z

The weak ATB condition requires every Coordinate daughter to contain a bindable
NP gap:

\ea
$\begin{aligned}
\ATB(c) \defas {} & \Coordination(c) \wedge
  (\forall k)[\Coordinate(c,k) \Rightarrow \CoordNPBindable(k)].
\end{aligned}$\\
\gloss{all Coordinate daughters contain an NP-bindable domain}
\z

That is not quite enough. It rules out the one-gap cases, but it does not yet
guarantee that the gaps are parallel in the right way. A stronger condition
quantifies over the actual vacancies:

\ea
$\begin{aligned}
\mathrm{ATB}^{+}(c) \defas {} & \Coordination(c) \wedge
  (\forall k)[\Coordinate(c,k) \Rightarrow
    (\exists z)[k \dm z \wedge \NPVac(z)]] \\
& {} \wedge
  (\forall k_1)(\forall k_2)(\forall z_1)(\forall z_2) \\
& \quad [\Coordinate(c,k_1) \wedge \Coordinate(c,k_2) \wedge
    k_1 \dm z_1 \wedge k_2 \dm z_2 \wedge {} \\
& \qquad \NPVac(z_1) \wedge \NPVac(z_2)
    \Rightarrow \ParallelGap(z_1,z_2)].
\end{aligned}$\\
\gloss{all Coordinate daughters contain NP vacancies, and those vacancies are syntactically parallel}
\z

\section*{4.~Worked instance}

In \mention{Those, I either sent back or paid for}, the relevant coordination
contains two VP coordinates. The first has an NP vacancy as complement of the
particle-verb complex \mention{sent back}; the second has an NP vacancy as
complement of \mention{for}. The coordinator \mention{or} is a Marker dependent
inside the final Coordinate, not itself a Coordinate. Each Coordinate daughter
therefore contains an NP-bindable domain, and the coordination satisfies the
weak $\ATB$ condition. If the two missing NPs are both linked to the same
prenucleus NP \mention{those}, they also satisfy the intended $\ParallelGap$
condition. This is a syntactic bridge condition, not a claim that the two gaps
have the same semantic role.

In \mention{Those, I either paid the bill or sent back}, the first Coordinate
daughter has no NP vacancy corresponding to \mention{those}. It therefore fails
$\ATB(c)$. The mirror-image example fails for the same reason in the second
Coordinate daughter.

\section*{5.~Open questions}

\begin{enumerate}[label=(\alph*),leftmargin=2em,itemsep=0.25em]
  \item $\ParallelGap$ is doing real work. If it cannot be reduced to ordinary
        function-labelled dominance plus the shared prenucleus relation, the
        fragment has merely hidden a SLASH-like dependency under a new name.
  \item Semantic or thematic role matching is deliberately absent. If ATB
        requires such a condition, it needs a separate interface predicate with
        its own warrant.
  \item The formula needs a sharper account of coordination structure:
        binary vs.\ flat coordination, unmarked coordinations, coordinators in
        Marker function inside the final Coordinate, markers such as
        \mention{either}, and possible non-constituent coordination.
  \item The weak condition may overgenerate cases with accidental gaps that
        are not jointly linked to the same prenucleus NP.
  \item This should be compared directly with Gazdar's 1981 treatment before
        being treated as a successful replacement.
\end{enumerate}

\end{document}
